A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2024-02-18 paper/paper.tex

First, we formulate the definition of valid renamings and prove that
semantic typing satisfies renaming. We say that $[[xi]]$ is
valid from the context $[[G]]$ to the context $[[D]]$ if the
following condition holds.
\[ \forall i \text{, if } i < | [[G]] |\text{, then }[[ ren xi i < |D| ]] \text{ and }
  [[D ren xi
  i < up Suc ren xi i > = G i < up Suc i > < xi >]] \]

@@

\section{Mechanization}
\label{sec:logrelmech}
\begin{figure}[h]
  TODO
  \caption{Statistics of the Coq Development}
  \label{fig:linecount}
\end{figure}

Figure~\ref{fig:linecount} shows the statics of our
development, including the base consistency proof from
Section~\ref{sec:logreldep} and \ref{sec:logrelproof} and the
extension from Section~\ref{sec:extension}.

>>>>

First, we formulate the definition of valid renamings and prove that
semantic typing satisfies renaming so we can weaken the context when
reasoning about the variable case of the fundamental lemma
(Lemma~\ref{lemma:stvar}). Intuitively, given a valuation $[[rho |= G
++ D]]$, it is easy to show that we can extract some valuation
$[[rho0]]$ such that $[[rho0 |= G]]$, where $[[rho0]]$ is obtained by
``truncating'' $[[rho]]$. As a result, if we know that $[[G |= a :
A]]$, then we can conclude that $[[G ++ D |= a0 : A0 ]]$, where
$[[a0]]$ and $[[A0]]$ are obtained by shifting $[[a]]$ and $[[A]]$
after weakening the context; this implication holds because $[[rho |=
G ++ D]]$ induces a context $[[rho0]]$ such that $[[rho0 |= G]]$ so we
can make use of the premise $[[G |= a : A]]$ to derive what we need
for the conclusion. We recommend the readers to skip the proofs of
Lemmas~\ref{lemma:validtruncate} through \ref{lemma:semrenaming}
during the first read as long as they have an intuitive understanding
of what the renaming property is meant to capture.

We say that $[[xi]]$ is
valid from the context $[[G]]$ to the context $[[D]]$ if the
following condition holds.
\[ \forall i \text{, if } i < | [[G]] |\text{, then }[[ ren xi i < |D| ]] \text{ and }
  [[D ren xi
  i < up Suc ren xi i > = G i < up Suc i > < xi >]] \]

@@

\section{Mechanization}
\label{sec:logrelmech}
\begin{figure}[h]
  \begin{tabular}{ c |  c  | c  }
    & Consistency & Normalization \\
    \hline
    Library (Autosubst 2)  & 491 & - \\
    Syntactic typing (specification) &  69 & - \\
    Renaming (common)  & 46 & -  \\
    Syntactic soundness & 660 &  - \\
    Untyped reduction & 350 & 834 \\
    Logical relation & 342 & 515 \\
    Semantic typing and soundness & 169 & 197 \\
  \end{tabular}
  \caption{Statistics of the Coq Development}
  \label{fig:linecount}
\end{figure}

Figure~\ref{fig:linecount} shows the statics of our
development, including the base consistency proof from
Section~\ref{sec:logreldep} and \ref{sec:logrelproof} and the
extension to normalization for open and closed terms from
Section~\ref{sec:extension}.

Since the normalization and the consistency development share the same
syntactic and typing specification, they differ in only the categories
that are related to the logical relation. We use - as a marker that
the line count is the same.

Autosubst 2 takes our syntax specification in higher-order abstract
syntax and generates Coq syntax specification, renaming and
substitution functions, and lemmas and tactics that allow reasoning
about those functions. The auto-generated files and the library header
files are counted toward the \emph{Library (Autosubst 2)} category.

Orthogonal to the development of our consistency and normalization
proof, we prove the syntactic soundness of our system through subject
reduction and progress. The syntactic soundness proof shares the
definition of renaming with our semantic soundness proof, which is
factored out as a separate file under the Renaming (common) category.
